Differences in score reliability do not explain meta-analytic heterogeneity in standardised effect sizes

### 2025-12-23 Manuscript_MD_ScoreReLiability.qmd
Update - FR comments

---
title: "Differences in score reliability do not explain meta-analytic heterogeneity in standardised effect sizes"
bibliography: ReLiability_references.bib
csl: styles/apa-6th-edition.csl
format: docx
editor: 
  markdown: 
    wrap: 72
---

@@

#### Non-technical abstract
Researchers often find that studies on the same psychological effect report different results, known as differences in effect sizes. Such variation suggests that the phenomenon is not yet fully understood. Some have argued that part of this variation comes from differences in measurement precision—how reliably a study’s tools or tests measure what they intend to measure. Based on this reasoning, it has been suggested that adjusting or “correcting” for these differences should make results more consistent across studies. We revisited this assumption by looking at large collaborative projects in psychology, such as ManyLabs, the Registered Replication Reports, and the Psychological Science Accelerator. We found that the expected reduction in variation does not generally occur. In fact, when measurement precision is already quite high - as it usually is in psychology, where researchers carefully select reliable measures - but samples differ widely in who participates, such corrections can actually increase differences in reported effect sizes. In simpler terms, differences in how diverse the samples of participants are across studies (which comes about from convenience sampling) matters more for explaining variation in results than small differences in how precisely a measure captures a construct. We introduce a simple rule that identifies when correcting for measurement differences is likely to backfire, increasing rather than reducing heterogeneity. We conclude by discussing why random measurement error plays a smaller role in explaining effect size differences than the use of diverse, convenience-based samples.

@@

One source for this heterogeneity, which is not a typical moderator tied
to a specific theory, might be impairments in score reliability which
vary in their extent across replications [@olsson2020]. Hunter and
Schmidt [-@hunter2004] and Wiernik and Dahlke [-@wiernik2020],
discussing the role of score reliability for standardized effect sizes,
claim that differences in score reliability should inflate, and
therefore increase, heterogeneity. This implies that standard
corrections for imperfect score reliability, such as attenuation
correction procedures, should remove that inflation, reducing
heterogeneity in corrected effect sizes.

Over the following pages, we assess whether score reliability can in
fact be made responsible for heterogeneity identified across several
psychological phenomena. However, we find no empirical evidence for the
argument that correcting for differences in score reliability reduces
effect size heterogeneity. Instead, to our great surprise, we find that
heterogeneity might as well be larger after the correction procedure.
Therefore, previous analyses and assumptions regarding the relationship
between score reliability and heterogeneity need to be false. In the
second half of this text we make use of analytical arguments to identify
the necessary conditions under which an attenuation correction
inevitably leads to an increase in effect size heterogeneity \[Do we
need any "Schleifen" in this paragraph - to reduce informational load /
density\].

@@

Previous discussions of heterogeneity in score reliability have
exclusively discussed it as a parameter that inflates heterogeneity in
ES, implying that, if score reliability was perfect across all
replications, the actual heterogeneity would have been lower. In their
discussion, Wiernik and Dahlke claim that „*Measurement error variance
will impact the results of meta-analyses in three ways: by (a) biasing
the mean effect size toward zero, (b) inflating effect-size
heterogeneity and confounding moderator effects, and (c) confounding
publication-bias and sensitivity analyses*” [p. 3, -@wiernik2020].
Additionally, more clearly, they state that “*If the studies included in
a meta-analysis differ in their measures’ reliabilities, heterogeneity
estimates will be artifactually inflated, erroneously suggesting larger
potential moderator effects*” [p. 4, -@wiernik2020]. They base their
claims largely on work done by Hunter and Schmidt [-@hunter2004], who
claim that “*Variation in reliability across studies causes variation in
the observed effect sizes above and beyond that produced by sampling
error.*” (p. 302). Overall, both references imply that differences in
score (un)reliability inevitably lead to an inflated heterogeneity in
ES.

@@

If information concerning score reliability is available, it is possible
to correct the individual ES for its unreliability. This process is also
known as attenuation correction and, while not without its criticisms
[@muchinsky1996; @schmidt1996; @schennach2016], is a repeatedly
recommended practice in Psychology [@hunter2004, @wiernik2020] . @eq-dc
describes a simple attenuation correction procedure.

@@

As all phenomena in the collection fulfil the first condition, 50
phenomena have been catalogued where the effect is studied by comparing
the outcome across two separate groups. From the collection of 50
phenomena, 21 fulfil the second condition, so that estimates of score
reliability can be derived. For the 21 phenomena, the dependent variable
used to construct the effect size is measured by employing more than a
single indicator. The number of indicators per scale varied between 3
and 30. The majority of scales was made up of 3 to 6 items. At the same
time, however, this implies that the remaining 29 phenomena did not make
use of more than a single indicator to measure the dependent variable.
Therefore, for the majority of phenomena it is not possible to assess in
how far measurement quality is sufficiently high in terms of e.g.
internal consistency or low random error variance.

@@

The estimates concerning the meta-analytic mean and heterogeneity of
these standardized effect sizes attempt to describe their population
distribution. Therefore, over the following paragraphs, whenever we
refer to estimates derived by means of random effects meta-analysis, we
will use the Greek letter $\delta_0$ to denote estimates of uncorrected
$d$'s population distribution and $\delta$ to denote estimates of
corrected $d_c$'s population distribution.

@@

Additionally, estimates of score reliability are derived. While not
without criticism [e.g. @sijtsma2009], Cronbach’s Alpha is used to
estimate score reliability. It is often noted that tau-equivalence is an
unrealistic assumption to hold against real-world data, implying that
other estimates of score reliability, such as McDonald’s Omega
[@mcdonald1999], Guttman’s Lambda 2 or 4 [@guttman1945], or the Greatest
Lower Bound [@tenBerge2004] might be better suited. However, all scales
employed in this project are scored by computing the simple mean or sum
of responses across items for each individual. Computing the mean or sum
of several items implicitly assumes tau-equivalence, as each item
contributes equally to the individual’s test score. Therefore,
Cronbach’s Alpha is the better choice to estimate score reliability, as
it avoids a mismatch in assumptions between how the score is computed
and how score reliability is estimated. The individual estimates of
Cohen’s $d$ are corrected using estimates of Cronbach's Alpha, according
to @eq-dc, including the corrected standard errors (@eq-SEdc). Thereby,
estimates of ES are computed, which are corrected for imperfect
reliability. Again, a random-effects meta-analysis is run using metafor,
generating an estimate of heterogeneity in corrected ES. Different
indicators of heterogeneity are readily available, such as $I^2$, $H^2$
or the coefficient of variation $CV$[@borenstein2009; @higgins2002].
However, the descriptions found in H&S and W&D discuss the absolute
amount of heterogeneity in terms of variance ($\tau^2$) or standard
deviation ($\tau$). Therefore, we also make use of absolute
heterogeneity here.

@@

[Equations @eq-d], [-@eq-SEd], [-@eq-dc], and [-@eq-SEdc] were used to
generate estimates of standardized ES, both uncorrected and corrected,
with their corresponding standard errors. Generally, this leads to
larger (absolute) ES and larger standard errors after the correction. As
an example from the 21 selected phenomena, the results from this
procedure on replications of Nosek’s study on sex differences in
preferences for the subject of Arts is displayed in the forest plot in
@fig-forest. Here, grey dots represent the uncorrected ES for each
sample, while black dots represent the corrected ES in each sample. The
bars surrounding these dots show the respective 95%-CIs.

@@

#### Which phenomena pass criterion 3)?

@tbl-meanES demonstrates that what we observed in @fig-forest holds
across all 21 phenomena. All meta-analytic effects are larger in their
absolute size after applying an attenuation correction procedure.
Similarly, the uncertainty quantified in the estimate's standard error
is larger. Additionally, @tbl-meanES highlights which phenomena pass
criterion (3)—the effect size must be statistically significantly
distinguishable from zero at $p < .05$. 12 phenomena pass this
criterion. It may be remarked that the attenuation correction procedure,
in this data, appears to have no influence on whether a phenomenon
passes the criterion. The differences in p-value are rather small and
lead to no difference in conclusion, regarding a significance level of
5%.

@@

*Note:* In this table $\hat{\mu}_{d}$ represents the meta-analytic
estimate of uncorrected $d$ and $\hat{\mu}_{d_c}$ represents the
estimate of the meta-analytic mean of corrected $d_c$.

@@

@tbl-tauES contains the results from performing an RG-MA on the
Cronbach's Alpha coefficients of the scales used to measure the 12
remaining phenomena. Additionally, it contains estimates of
heterogeneity for both uncorrected ES ($d$) and corrected ES ($d_c$),
which are central to assess the research question posed in section
*Attenuation Correction*. Most importantly, @tbl-tauES demonstrates that
for all but one of the phenomena, the attenuation correction procedure
barely has an impact on the extent of ES heterogeneity. However, while
the heterogeneity for the phenomenon from RRR9 (Behaviour) seems to be
reduced by about a third of its initial value, the ES heterogeneity for
this phenomenon is rather small to begin with. Using a significance
test, we can not distinguish this value confidently from zero. This
implies that for the majority of phenomena, differences in score
reliability can not be made responsible for heterogeneity in
standardized effect sizes.

@@

gt::gt(tab_both  %>% 
  arrange(tab_hetES.MASC)) %>% 
  gt::cols_label(tab_hetES.MASC = "MASC",
                 mu_alpha_str = gt::md("$\\hat{\\mu}_{\\rho_{XX'}}$"),
                 tau_alpha = gt::md("$\\hat{\\tau}_{\\rho_{XX'}}$"),
                 QE_str = "QE (df)",
                 QEp = "p",
                 tau_raw = gt::md("$\\hat{\\tau}_{d}$"),
                 QE_raw_str = "QE (df)",
                 QEp_raw = "p",
                 tau_raw = gt::md("$\\hat{\\tau}_{d_c}$"),
                 QE_raw_str = "QE (df)",
                 QEp_raw = "p") %>% 
  gt::tab_options(table.width = gt::pct(100),
                  column_labels.font.weight = "bold") #%>% 
  # gt::cols_width(pval_raw ~ gt::px(40),
  #                pval_cor ~ gt::px(40))

@@

*Note:* $\hat{\mu}_{\rho_{XX'}}$ is the back-transformed meta-analytic
estimate of mean score reliability; $\tau_{\mu_{\rho_{XX'}}}$ is the
back-transformed estimate of heterogeneity in score reliability;
$\hat{\tau}_{d}$ represents the estimate of heterogeneity in uncorrected
$d$; $\hat{\tau}_{d_c}$ represents the estimate of the heterogeneity in
corrected $d_c$.

Secondly, in @tbl-tauES we find that for ten out of twelve phenomena, ES
heterogeneity grows as a result from the attenuation correction
procedure. However, it needs to be noted that these differences are
small and could not be distinguished from sampling uncertainty in a
statistical test. For this reason, we do not mean to interpret these
differences, while we note that the direction of differences are clearly
inconsistent with expectations formulated in the literature discussed so
far. These results are incompatible with claims made in W&D and H&S.

@@

Concerning the remaining two phenomena, for the phenomenon described in
ML5 - Shnabel (RPP), the reduction in heterogeneity is miniscule and the
estimate of heterogeneity is not statistically significant in the first
place. Most likely, the difference in ES heterogeneity estimates can be
explained by issues of estimation, as only eight replications are
available. Similarly, only for the phenomenon described in RRR9
(Behaviour), we find a seemingly substantial reduction in heterogeneity
due to the attenuation correction procedure. The estimates of ES
heterogeneity are not statistically significant, however, which hampers
our ability to meaningfully interpret and compare these estimates. For
the majority of phenomena it appears that differences in score
reliability do not inflate heterogeneity in uncorrected effect sizes.
Instead these differences mask ES heterogeneity, that can be unveiled by
accounting for score reliability.

@@

It might help to think about linear models, when discussing the
intuitive idea that controlling for score reliability should reduce ES
heterogeneity. Heterogeneity describes unexplained variation in a
parameter's true distribution, that we model using meta-analytical
techniques [@borenstein2009; @higgins2002]. In that way, the idea of
heterogeneity is closely related to residual variance, as we observe it
in a simple linear model. Both describe variation that could not be
explained by the model specifications, be it a model that employs
predictive variables like a simple linear model or a meta-analytic model
incorporating uncertainty in terms of standard errors.

The residual variance that remains in a simple linear model will be
reduced in all instances, where we use a predictor that we know shares
variance with the predicted parameter. Generally, removing a source of
variance by controlling for it, we expect any unexplained variation to
shrink. Similarly, we know that differences in score reliability, by
definition, are related to differences in standardized effect sizes.
Therefore, intuitively, it seems sensible to expect that controlling for
differences in score reliability should reduce the unexplained variance
in standardized effect sizes, as described in H&S and W&D.

Hunter & Schmidt have additionally derived an equation which
demonstrates how heterogeneity in score reliability affects
heterogeneity in ESs [-@hunter2004, p. 309]. @eq-HS implies that
heterogeneity in $\delta_0$ is essentially a function in which
heterogeneity and mean value of $\delta$ are weighted by mean score
reliability $\rho_{xx‘}$ and its heterogeneity.

$$\tau^2_{\delta_0} = [\mu_{\rho_{xx‘}}]^2 \tau^2_{\delta} + [\mu_{\delta}]^2 \tau^2_{\rho_{xx‘}}$$ {#eq-HS}

We adjusted @eq-HS using the notation used throughout this text. Here,
$\delta_0$, carrying a 0 in the index, refers to the uncorrected ES,
undistorted by sampling error but not corrected for measuring error.
Without an index, $\delta$, on the other hand, refers to the corrected
ES, undistorted by sampling error and measurement error. Lastly,
$\rho_{XX'}$ refers to score reliability. Hunter & Schmidt
[-@hunter2004] claim that @eq-HS demonstrates that differences in score
reliability inflate ES heterogeneity, which means that an attenuation
correction procedure would have to reduce ES heterogeneity. However, it
can be easily verified that @eq-HS does not allow for such claims.
Depending on both the mean value $\mu_{\rho_{xx‘}}$ and the
heterogeneity $\tau^2_{\rho_{xx‘}}$ in score reliability, ES
heterogeneity might be increased or decreased by the differences in
score reliability.

Most likely, Hunter and Schmidt did not fully explore the implications
of @eq-HS, but followed the general intuition we lined out–"If a source
of variation is controlled for, the amount of unexplained variance
should be reduced." Wiernik and Dahlke [-@wiernik2020] refer to the work
by Hunter & Schmidt [-@wiernik2020], likely following their intuition.

In the re-analysis of archival data, we have demonstrated that this
intuition does not hold for the relationship of ES heterogeneity and
score reliability. Instead, we propose an alternative way that helps us
discern, under which conditions we can expect an attenuation correction
procedure to lead to an increase in ES heterogeneity.

@@

Realising that an ES, as defined in @eq-d, is actually a ratio, allows
for a different analytical description of how differences in score
reliability can affect ES heterogeneity. If the phenomenon is
heterogeneous and score reliability varies across replications, it seems
sensible to assume that both numerator and denominator (MD and
$\sigma_X$) from @eq-d vary across replications. In that case, the ES
can be described as a random variable stemming from a ratio
distribution. A ratio distribution is a probability distribution
constructed by dividing one random variable by a second random variable:
$Z = \frac{X}{Y}$. First order taylor approximation may be used to
generate estimates of mean and variance of the ratio variable
[@vanKempen1999]. Taylor approximations typically require no specific
assumptions regarding the distribution form, but that the variable (here
a ratio) is a) differentiable, b) has existing moments and that c) its
higher-order moments are negligible (e.g. limited variability and
monotonicity) (Asymptotic statistics, A.W. van der Vaart - find book
online). Here we use the Taylor-estimator of the ratio’s variance to
demonstrate how differences in score reliability may affect
heterogeneity in ES.

@@

$$\tau^2[\delta_0] \approx \frac{\tau^2[MD]}{\mu[\sigma_x]^2} + \frac{\mu[MD]^2}{\mu[\sigma_x]^4} \tau^2[\sigma_x]$$ {#eq-ratio_d0}

@@

In that case, some heterogeneity in $\sigma_T$ would remain, albeit
heterogeneity found in $\sigma_x$ was reduced. @eq-ratio_d0 essentially
demonstrates how mean and heterogeneity of MD, paired with mean and
heterogeneity of total score standard deviation $\sigma_x$ can be used
to estimate heterogeneity in raw ES $\delta_0$. Similarly, we can adjust
@eq-ratio_d0 to describe how, instead of total score standard deviation,
mean and variance in true score standard deviation $\sigma_T$ affect
heterogeneity in corrected ES $\delta$.

$$\tau^2[\delta] \approx \frac{\tau^2[MD]}{\mu[\sigma_T]^2} + \frac{\mu[MD]^2}{\mu[\sigma_T]^4} \tau^2[\sigma_T]$$ {#eq-ratio_d}

The differences between the two [Equations @eq-ratio_d0] and
[-@eq-ratio_d] demonstrates how ES heterogeneity is affected by an
attenuation correction procedure. According to the quotes in H&S and
W&D, @eq-ratio_d0 should lead to a larger value of
$\tau^2\left[\delta_0\right]$, compared to the estimate of
$\tau^2\left[\delta\right]$ from @eq-ratio_d. However, as demonstrated
in the re-analysis of archival data, we can already expect that this can
not be an implication of these Equations.

@@

$$\tau^2_{\delta} > \tau^2_{\delta_0} \quad  \quad \quad \quad \quad  \frac{\tau_{MD}^2}{\mu_{\sigma_T}^2} + \frac{\mu_{MD}^2}{\mu_{\sigma_T}^4} \tau^2_{\sigma_T} > \frac{\tau_{MD}^2}{\mu_{\sigma_x}^2} + \frac{\mu_{MD}^2}{\mu_{\sigma_x}^4} \tau^2_{\sigma_x}$$ {#eq-ineq_init}

@@

In our example of $R_1 = .8$, the true score variance is 4 times as large
as the error score variance. However, for the attenuation correction to
reduce ES heterogeneity, the smaller error score variance would need to
account for almost 50 % of heterogeneity in score variance. This asymmetry
grows even larger, as the average score reliability grows larger.
This demonstrated that the individual size of either
mean score variance component or its heterogeneity is not of importance.
Instead, the ratio of these components determine, whether heterogeneity
can decrease after an attenuation correction.

>>>>

---
title: "Differences in score reliability do not explain meta-analytic heterogeneity in standardised effect sizes"
bibliography: ReLiability_references.bib
csl: styles/apa-7th-edition.csl
format: docx
editor: 
  markdown: 
    wrap: 72
---

@@

#### Non-technical abstract

Researchers often find that studies on the same psychological effect
report different results, known as differences in effect sizes. Such
variation suggests that the phenomenon is not yet fully understood. Some
have argued that part of this variation comes from differences in
measurement precision—how reliably a study’s tools or tests measure what
they intend to measure. Based on this reasoning, it has been suggested
that adjusting or “correcting” for these differences should make results
more consistent across studies. We revisited this assumption by looking
at large collaborative projects in psychology, such as ManyLabs, the
Registered Replication Reports, and the Psychological Science
Accelerator. We found that the expected reduction in variation does not
generally occur. In fact, when measurement precision is already quite
high - as it usually is in psychology, where researchers carefully
select reliable measures - but samples differ widely in who
participates, such corrections can actually increase differences in
reported effect sizes. In simpler terms, differences in how diverse the
samples of participants are across studies (which comes about from
convenience sampling) matters more for explaining variation in results
than small differences in how precisely a measure captures a construct.
We introduce a simple rule that identifies when correcting for
measurement differences is likely to increase rather than reduce
heterogeneity. We conclude by discussing why random measurement error
plays a smaller role in explaining effect size differences than the use
of diverse, convenience-based samples.

@@

The search for moderators that might explain said heterogeneity across
several phenomena has not been particularly successful [e.g. @ManyLabs3,
@ManyLabs2]. Another source for this heterogeneity, which is not a
typical moderator tied to a specific theory, might be impairments in
score reliability which vary in their extent across replications
[@olsson2020]. Hunter and Schmidt [-@hunter2004] and Wiernik and Dahlke
[-@wiernik2020], discussing the role of score reliability for
standardized effect sizes, claim that differences in score reliability
should inflate, and therefore increase, heterogeneity. This implies that
standard corrections for imperfect score reliability, such as
attenuation correction procedures, should remove that inflation,
reducing heterogeneity in corrected effect sizes.

Over the following pages, we assess whether score reliability can in
fact be made responsible for heterogeneity identified across several
psychological phenomena. However, we find no empirical evidence for the
argument that correcting for differences in score reliability reduces
effect size heterogeneity. Instead, we find that heterogeneity might as
well be larger after the correction procedure. Therefore, previous
analyses and assumptions regarding the relationship between score
reliability and heterogeneity need to be false. In the second half of
this text we make use of analytical arguments to identify the necessary
conditions under which an attenuation correction inevitably leads to an
increase in effect size heterogeneity.

@@

Previous discussions of heterogeneity in score reliability have
exclusively discussed it as a parameter that inflates heterogeneity in
ES, implying that, if score reliability was perfect across all
replications, the actual heterogeneity would have been lower. In their
discussion, Wiernik and Dahlke claim that „*Measurement error variance
will impact the results of meta-analyses in three ways: by (a) biasing
the mean effect size toward zero, (b) inflating effect-size
heterogeneity and confounding moderator effects, and (c) confounding
publication-bias and sensitivity analyses*” [p. 3, -@wiernik2020].
Additionally, more clearly, they state that “*If the studies included in
a meta-analysis differ in their measures’ reliabilities, heterogeneity
estimates will be artifactually inflated, erroneously suggesting larger
potential moderator effects*” [p. 4, -@wiernik2020]. They base their
claims largely on work done by Hunter and Schmidt [-@hunter2004], who
claim that “*Variation in reliability across studies causes variation in
the observed effect sizes above and beyond that produced by sampling
error.*” (p. 302). Overall, both references imply that differences in
score reliability inevitably lead to an inflated heterogeneity in ES.

@@

If information concerning score reliability is available, it is possible
to correct the individual ES for its unreliability. This process is also
known as attenuation correction and, while not without its criticisms
[@muchinsky1996; @schmidt1996; @schennach2016], is a repeatedly
recommended practice in Psychology [@hunter2004; @wiernik2020]. @eq-dc
describes a simple attenuation correction procedure.

@@

As all phenomena in the collection fulfil the first condition, 50
phenomena have been catalogued where the effect is studied by comparing
the outcome across two separate groups. From the collection of 50
phenomena, 21 fulfil the second condition, so that estimates of score
reliability can be derived. For the 21 phenomena, the dependent variable
used to construct the effect size is measured by employing more than a
single indicator. The number of indicators per scale varied between 3
and 30. The majority of scales was made up of 3 to 6 items. At the same
time, however, this implies that the remaining 29 phenomena did not make
use of more than a single indicator to measure the dependent variable.
Therefore, for the majority of phenomena it is not possible to assess
whether measurement quality is sufficiently high in terms of internal
consistency or low random error variance.

@@

The estimates concerning the meta-analytic mean and heterogeneity of
these standardized effect sizes attempt to describe their population
distribution. Therefore, over the following paragraphs, whenever we
refer to estimates derived by means of random effects meta-analysis, we
will use the Greek letter $\delta$ to denote estimates of uncorrected
$d$'s population distribution and $\delta_c$ to denote estimates of
corrected $d_c$'s population distribution.

@@

Additionally, estimates of score reliability are derived. While not
without criticism [e.g. @sijtsma2009], Cronbach’s Alpha is used to
estimate score reliability. It is often noted that tau-equivalence is an
unrealistic assumption to hold against real-world data, implying that
other estimates of score reliability, such as McDonald’s Omega
[@mcdonald1999], Guttman’s Lambda 2 or 4 [@guttman1945], or the Greatest
Lower Bound [@tenBerge2004] might be better suited. However, all scales
employed in this project are scored by computing the simple mean or sum
of responses across items for each individual. Computing the mean or sum
of several items implicitly assumes tau-equivalence, as each item
contributes equally to the individual’s test score. Therefore,
Cronbach’s Alpha is the better choice to estimate score reliability, as
it avoids a mismatch in assumptions between how the score is computed
and how score reliability is estimated. The individual estimates of
Cohen’s $d$ are corrected using estimates of Cronbach's Alpha, according
to @eq-dc, including the corrected standard errors (@eq-SEdc). Thereby,
estimates of ES are computed, which are corrected for imperfect
reliability. Again, a random-effects meta-analysis is run using metafor,
generating an estimate of heterogeneity in corrected ES. Different
indicators of heterogeneity are readily available, such as $I^2$, $H^2$
[@borenstein2009; @higgins2002]. A relative heterogeneity measure less
common in Psychology, but easily computed is the coefficient of
variation $CV = \frac{\tau}{\mu}$. However, the descriptions found in
H&S and W&D discuss the absolute amount of heterogeneity in terms of
variance ($\tau^2$) or standard deviation ($\tau$). Therefore, we also
make use of absolute heterogeneity here.

@@

[Equations @eq-d], [-@eq-dc], [-@eq-SEd], and [-@eq-SEdc] were used to
generate estimates of standardized ES, both uncorrected and corrected,
with their corresponding standard errors. Generally, this leads to
larger (absolute) ES and larger standard errors after the correction. As
an example from the 21 selected phenomena, the results from this
procedure on replications of Nosek’s study on sex differences in
preferences for the subject of Arts is displayed in the forest plot in
@fig-forest. Here, grey dots represent the uncorrected ES for each
sample, while black dots represent the corrected ES in each sample. The
bars surrounding these dots show the respective 95%-CIs.

@@

#### Which phenomena pass criterion (3)?

@tbl-meanES demonstrates that what we observed in @fig-forest holds
across all 21 phenomena. All meta-analytic effects are larger in their
absolute size after applying an attenuation correction procedure.
Similarly, the uncertainty quantified in the estimate's standard error
is larger. Additionally, @tbl-meanES highlights which phenomena pass
criterion (3)—the effect size must be statistically significantly
distinguishable from zero at $p < .05$. Twelve phenomena pass this
criterion. It may be remarked that the attenuation correction procedure,
in this data, appears to have no influence on whether a phenomenon
passes the criterion. The differences in p-value are rather small and
lead to no difference in conclusion, regarding a significance level of
5%.

@@

*Note:* In this table $\hat{\mu}_{d}$ represents the meta-analytic
estimate of uncorrected $d$ and $\hat{\mu}_{d_c}$ represents the the
meta-analytic estimate of corrected $d_c$.

@@

@tbl-tauES contains the results from performing an RG-MA on the
Cronbach's Alpha coefficients of the scales used to measure the 12
remaining phenomena. Additionally, it contains estimates of
heterogeneity for both uncorrected ES ($d$) and corrected ES ($d_c$),
which are central to assess the research question posed in section
*Attenuation Correction*. Most importantly, @tbl-tauES demonstrates that
for all but one of the phenomena, the attenuation correction procedure
barely has an impact on the extent of ES heterogeneity. However, while
the heterogeneity for the phenomenon from RRR9 (Behaviour) seems to be
reduced by about a third of its initial value, the ES heterogeneity for
this phenomenon is rather small to begin with. Using a significance
test, we can not distinguish this value confidently from zero. This
implies that for none of phenomena, differences in score reliability can
be made responsible for heterogeneity in standardized effect sizes.

@@

gt::gt(tab_both  %>% 
  arrange(tab_hetES.MASC)) %>% 
  gt::cols_label(tab_hetES.MASC = "MASC",
                 mu_alpha_str = gt::md("$\\hat{\\mu}_{\\rho_{XX'}}$"),
                 tau_alpha = gt::md("$\\hat{\\tau}_{\\rho_{XX'}}$"),
                 QE_str = "QE (df)",
                 QEp = "p",
                 tau_raw = gt::md("$\\hat{\\tau}_{d}$"),
                 QE_raw_str = "QE (df)",
                 QEp_raw = "p",
                 tau_cor = gt::md("$\\hat{\\tau}_{d_c}$"),
                 QE_cor_str = "QE (df)",
                 QEp_cor = "p") %>% 
  gt::tab_options(table.width = gt::pct(100),
                  column_labels.font.weight = "bold") #%>% 
  # gt::cols_width(pval_raw ~ gt::px(40),
  #                pval_cor ~ gt::px(40))

@@

*Note:* $\hat{\mu}_{\rho_{XX'}}$ is the back-transformed meta-analytic
estimate of mean score reliability; $\hat{\tau}_{\mu_{\rho_{XX'}}}$ is
the back-transformed estimate of heterogeneity in score reliability;
$\hat{\tau}_{d}$ represents the estimate of heterogeneity in uncorrected
$d$; $\hat{\tau}_{d_c}$ represents the estimate of the heterogeneity in
corrected $d_c$.

Secondly, in @tbl-tauES we find that for ten out of twelve phenomena, ES
heterogeneity grows as a result from the attenuation correction
procedure. However, it needs to be noted that these differences are
rather small. For this reason, we do not mean to interpret these
differences, while we note that the direction of differences are clearly
inconsistent with expectations formulated in the literature discussed so
far. These results are incompatible with claims made in W&D and H&S.

@@

Concerning the remaining two phenomena, for the phenomenon described in
ML5 - Shnabel (RPP), the reduction in heterogeneity is miniscule and the
estimate of heterogeneity is not statistically significant in the first
place. Most likely, the difference in ES heterogeneity estimates can be
explained by issues of estimation, as only eight replications are
available. Similarly, only for the phenomenon described in RRR9
(Behaviour), we find a seemingly substantial reduction in heterogeneity
due to the attenuation correction procedure. The estimates of ES
heterogeneity are not statistically significant, however, which hampers
our ability to meaningfully interpret and compare these estimates. For
the majority of phenomena it appears that differences in score
reliability do not inflate heterogeneity in uncorrected effect sizes.
Instead these differences mask ES heterogeneity that can be unveiled by
accounting for score reliability.

@@

It might help to think about linear models, when discussing the
intuitive idea that controlling for score reliability should reduce ES
heterogeneity. Heterogeneity describes unexplained variation in a
parameter's true distribution, that we model using meta-analytic
techniques [@borenstein2009; @higgins2002]. In that way, the idea of
heterogeneity is closely related to residual variance, as we observe it
in a simple linear model. Both describe variation that could not be
explained by the model specifications, be it a model that employs
predictive variables like a simple linear model or a meta-analytic model
incorporating uncertainty in terms of standard errors.

The residual variance that remains in a simple linear model is reduced
in all instances, where we use a predictor that we know shares variance
with the predicted parameter. Generally, removing a source of variance
by controlling for it, we expect any unexplained variation to shrink.
Similarly, we know that differences in score reliability, by definition,
are related to differences in standardized effect sizes. Therefore,
intuitively, it seems sensible to expect that controlling for
differences in score reliability should reduce the unexplained variance
in standardized effect sizes, as described in H&S and W&D.

Hunter & Schmidt have additionally derived an equation which
demonstrates how heterogeneity in score reliability affects
heterogeneity in ESs [-@hunter2004, p. 309]. @eq-HS implies that
heterogeneity in $\delta$ is essentially a function in which
heterogeneity and mean value of $\delta_c$ are weighted by mean score
reliability $\rho_{xx‘}$ and its heterogeneity. Here, we adjusted the
notation from Hunter & Schmidt [-@hunter2004] to the notation used
through this text.

$$\tau^2_{\delta} = [\mu_{\rho_{xx‘}}]^2 \tau^2_{\delta_c} + [\mu_{\delta_c}]^2 \tau^2_{\rho_{xx‘}}$$ {#eq-HS}

Here, $\delta$ refers to the uncorrected ES, undistorted by sampling
error but not corrected for measuring error. With the index $c$,
$\delta_c$, on the other hand, refers to the corrected ES, undistorted
by sampling error and measurement error. Lastly, $\rho_{XX'}$ refers to
score reliability. Hunter & Schmidt [-@hunter2004] claim that @eq-HS
demonstrates that differences in score reliability inflate ES
heterogeneity, which means that an attenuation correction procedure
would have to reduce ES heterogeneity. However, it can be easily
verified that @eq-HS does not allow for such claims. Depending on both
the mean value $\mu_{\rho_{xx‘}}$ and the heterogeneity
$\tau^2_{\rho_{xx‘}}$ in score reliability, ES heterogeneity might be
increased or decreased by the differences in score reliability.

Most likely, Hunter and Schmidt did not fully explore the implications
of @eq-HS, but followed the general intuition we lined out – "If a
source of variation is controlled for, the amount of unexplained
variance should be reduced." Wiernik and Dahlke [-@wiernik2020] refer to
the work by Hunter & Schmidt [-@hunter2004], likely following their
intuition.

In the re-analysis of archival data, we have demonstrated that this
intuition does not hold for the relationship of ES heterogeneity and
score reliability. Instead, we propose an alternative way that helps us
discern under which conditions we can expect an attenuation correction
procedure to lead to an increase in ES heterogeneity.

@@

Realising that an ES, as defined in @eq-d, is actually a ratio, allows
for an analytical description alternative to @eq-HS of how differences
in score reliability can affect ES heterogeneity. If the phenomenon is
heterogeneous and score reliability varies across replications, it seems
sensible to assume that both numerator and denominator (MD and
$\sigma_X$) from @eq-d vary across replications. In that case, the ES
can be described as a random variable stemming from a ratio
distribution. A ratio distribution is a probability distribution
constructed by dividing one random variable by a second random variable:
$Z = \frac{X}{Y}$. First order Taylor approximation may be used to
generate estimates of mean and variance of the ratio variable
[@vanKempen1999]. Taylor approximations typically require no specific
assumptions regarding the distribution form, but that the variable (here
a ratio) is a) differentiable, b) has existing moments and that c) its
higher-order moments are negligible [e.g. @vandervaart2000]. Here we use
the Taylor-estimator of the ratio’s variance to demonstrate how
differences in score reliability may affect heterogeneity in ES.

@@

$$\tau^2[\delta] \approx \frac{\tau^2[MD]}{\mu[\sigma_x]^2} + \frac{\mu[MD]^2}{\mu[\sigma_x]^4} \tau^2[\sigma_x]$$ {#eq-ratio_d0}

@@

In that case, some heterogeneity in $\sigma_T$ would remain, albeit
heterogeneity found in $\sigma_X$ was reduced. @eq-ratio_d0 essentially
demonstrates how mean and heterogeneity of MD, paired with mean and
heterogeneity of total score standard deviation $\sigma_X$ can be used
to estimate heterogeneity in raw ES $\delta$. Similarly, we can adjust
@eq-ratio_d0 to describe how, instead of total score standard deviation,
mean and variance in true score standard deviation $\sigma_T$ affect
heterogeneity in corrected ES $\delta_c$.

$$\tau^2[\delta_c] \approx \frac{\tau^2[MD]}{\mu[\sigma_T]^2} + \frac{\mu[MD]^2}{\mu[\sigma_T]^4} \tau^2[\sigma_T]$$ {#eq-ratio_d}

The differences between the two [Equations @eq-ratio_d0] and
[-@eq-ratio_d] demonstrate how ES heterogeneity is affected by an
attenuation correction procedure. According to the quotes in H&S and
W&D, @eq-ratio_d0 should lead to a larger value of
$\tau^2\left[\delta\right]$, compared to the estimate of
$\tau^2\left[\delta_c\right]$ from @eq-ratio_d. However, as demonstrated
in the re-analysis of archival data, we can already expect that this can
not be an implication of these Equations.

@@

$$\tau^2_{\delta_c} > \tau^2_{\delta} \quad  \quad \quad \quad \quad  \frac{\tau_{MD}^2}{\mu_{\sigma_T}^2} + \frac{\mu_{MD}^2}{\mu_{\sigma_T}^4} \tau^2_{\sigma_T} > \frac{\tau_{MD}^2}{\mu_{\sigma_x}^2} + \frac{\mu_{MD}^2}{\mu_{\sigma_x}^4} \tau^2_{\sigma_x}$$ {#eq-ineq_init}

@@

In our example of $R_1 = .8$, the true score variance is 4 times as
large as the error score variance. However, for the attenuation
correction to reduce ES heterogeneity, the smaller error score variance
would need to account for almost 50 % of heterogeneity in score
variance. This asymmetry grows even larger, as the average score
reliability grows larger. This demonstrates that the individual size of
either mean score variance component or its heterogeneity is not of
importance. Instead, the ratio of these components determine whether
heterogeneity can decrease after an attenuation correction.
